MASCARA: coexpression analysis in data from designed experiments

### 2024-02-08 R/Paper_markdown.Rmd
24_02_08_c

# Fred T.G. White^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands, Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands *], Anna Heintz-Buschart^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Lemeng Dong^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Harro J. Bouwmeester^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Johan A. Westerhuis^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Age K. Smilde^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands]

# - name: Fred T.G. White
#     email: f.t.g.white@uva.nl
#     affiliation: 
#       - University of Amsterdam
#       - Another example
#     corresponding: alice@example.com
#   - name: Bob Security
#     email: bob@example.com
#     affiliation: 
#       - Another University
#       - Some Institute of Technology
# address:
#   - code: Univeristy of Amsterdam
#     address: Department 1, Street, City, State, Zip
#   - code: Another example
#     address: Department 2, Street, City, State, Zip

## CHANGE THIS   
#"one comment already, since I am looking at it right now. In the readme “~/MASCARA/R” , the repo (clone) most likely doesn’t live there. Maybe start from ‘clone this repo’ and then let people set the wd to where they put the repo. Or put the snippet from the readme into an r script that you let them open in rstudio to make sure they’re in the right place" AHB
```

@@

<!-- For the rest of the manuscript we take the strigolactones as our example biological pathway. -->

@@

#CHANGE caption

@@

<!-- \onecolumn ### -->

@@

<!-- \Begin{multicols}{2} -->

@@

<!-- 1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \). -->
<!-- 2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \): -->
<!--    a. Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively. -->
<!--    b. Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \). -->

@@

<!-- and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$ -->
<!-- PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. -->

@@

# Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")  
Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")

@@

<!-- The results from above were obtained from datasets with varying noise, with core aspects such as the underlying structure, the number of replicates, the strength of correlation between variables, and the size of the noise remained fixed. With the following, we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely: -->

<!-- 1. Within group correlation structure -->
<!-- 2. Between group correlation structure -->
<!-- 3. Random noise -->

<!-- Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size \@ref(fig:simOverview)) across all iterations of this test. We vary the sizes of the other two variance parameters (*d* and *l*). Furthermore we also show the effect of replicate number and demonstrate that MASCARA (along with the other methods) can be applied to data with generic profiles, not limited to the specific structures that were used to demonstrate the differences in coexpression analysis above. Here we create 20 datasets with random combined effect and within group correlation structures. -->

```{r, echo = F}
#30 random initial scores, similar noise 
#parameters, 3-30 replicates per condition. 
# Rep_Tests <- readRDS("Noise_Sim_23_07_29_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS

Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS

@@

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method, fill = Method)) +
  scale_color_viridis(discrete = T)+
  # scale_fill_viridis(discrete = T)+

        # scale_shape_manual(values = c(0:3)) +
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_dark() +
      theme(axis.title = element_blank())

@@

Taken together our results indicate that the lack of ability to discern different types of variance in a dataset can compromise the detection of true coexpression patterns. MASCARA accounts for the different types of (experimental and ambient) variance and enables finer scale detection of within group variance structure.

@@

```{r}
Rep_Tests <- readRDS("Noise_Sim_23_07_30_FS_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
names(Rep_Tests) <- c("TP","SR","VIP","SVD1")

>>>>

@@

@@

@@

@@

@@

@@

@@

Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")

@@

```{r, echo = F}
#30 random initial scores, similar noise 
#parameters, 3-30 replicates per condition.

Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")

@@

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method, fill = Method)) +
  scale_color_viridis(discrete = T)+
    stat_summary() +
    ylim(0,max ) +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  theme_dark() +
      theme(axis.title = element_blank())

@@

Taken together our results indicate that the lack of ability to discern different types of variance in a dataset can compromise the detection of true coexpression patterns. MASCARA accounts for the different types of (experimental and ambient) variance and enables finer scale detection of within group variance structure. Code and data for this project are available at https://github.com/BiosystemsDataAnalysis/MASCARA.

@@

```{r fig.cap="Feature selection simulation, selectivity ratio; SR, target projection; TP and variable importance in projection (VIP) compareed at 3 levels of noise, indicated in facet titles."}
Rep_Tests <- readRDS("Noise_Sim_23_07_30_FS_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
names(Rep_Tests) <- c("TP","SR","VIP","SVD1")
